How to build an homogeneous language for heterogeneous platforms

### 2013-05-24 papers/gpce2013/gpce2013.tex
plop

\begin{abstract}
Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
applications more tedious. We use a deep embedded DSL to define a common language abstracting over platform
differences for the developers point of view, but generating efficient code leveraging the target platforms
characteristics and giving them control on the amount of code produced on client-side.
\end{abstract}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintenance effort: if a piece of logic was implemented on
client-side and it turns out that it needs to be implemented on server-side instead, the code can not be reused and
the feature needs to be completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both
client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
when a page is requested by a client, but they may also be built from the client-side to perform an incremental
update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
both client-side and server-side? The more interactive the application is, the more logic needs to be duplicated
between the server-side and the client-side (explain why?).

(Mention the problem of using client-side libraries to get high-level abstractions: double runtime overhead --
download and execution)

Using the same programming language on both server-side and client-side can improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
action language natively supported by most of Web clients -- can be used on server-side, and an increasing
number of programming languages or compiler back-ends can generate JavaScript code (\eg
Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

However, using the same programming language is not enough because the client and server programming environments
are not the same. For instance, DOM fragments can be defined on client-side using the standard DOM API, but this
API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make it leverage
the native APIs, when possible, for performance reasons?

@@

\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is a compiled embedded DSL defined in Scala that generates JavaScript
code. In this paper we present additional contributions on top of this DSL. We use staging (explain) to generate
efficient code for typical abstractions used for Web programming and to generate efficient specialized code for both
client and server sides for shared abstractions. More precisely, we demonstrate the following features:

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
 \item (Modules?)
\end{itemize}

(More on the results metrics)

>>>>

\begin{abstract}
Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
applications tedious. We use a deep embedded DSL to define a common language abstracting over platform differences
for the developers point of view, but generating efficient code leveraging the target platforms characteristics and
giving them control on the amount of code produced on client-side.
\end{abstract}

@@

Web applications are attractive because they require no installation or deployment steps on clients and enable large
scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
engineering process and requires a higher maintenance effort: if a piece of logic is implemented on client-side and
finally needs to be implemented on server-side instead, the code can not be reused and the feature needs to be
completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both client-side and server-side
need to be duplicated. For instance, HTML fragments may be built from the server-side when a page is requested by a
client, but they may also be built from the client-side to perform an incremental update subsequent to an user
action. How could developers write HTML fragment definitions once and render them on both client-side and
server-side?

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
frameworks to get higher-level abstractions on client-side, making their code easier to reason about and to
maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
-- has no static linking system allowing to pick just the functions actually used by a program, forcing clients to
download the whole library even though only a few functions are actually used.

Using the same programming language on both server-side and client-side can improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language can be used on server-side,
and an increasing number of programming languages or compiler back-ends can generate JavaScript code (\eg
Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

However, using the same programming language is not enough because the client and server programming environments
are not the same. For instance, DOM fragments can be defined on client-side using the standard DOM API, but this
API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make the
executable code leverage the native APIs, when possible, for performance reasons?

@@

\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is such a compiled embedded DSL defined in Scala that generates
JavaScript code, making it possible to write the client-side code of Web applications using JavaScript. This paper
enriches js-scala in order to solve the problems described above. We use staging (explain) to generate efficient code
for typical abstractions used for Web programming and to generate efficient specialized code for both client and
server sides for shared abstractions. More precisely, we demonstrate the following features:

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Static linking.
\end{itemize}

Compared to libraries offering the same level of abstraction, we observe that js-scala produces significantly smaller
executables with better runtime performances.
